Global code motion of congruent instructions

### 2017-04-17 paper.tex
perf data

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  We present a global code motion compiler optimization which schedules
identical computations across the program so as to save code size.  Not only
this code motion saves code size, it exposes redundancies in some cases, it
exposes more instruction level parallelism in the basic block when the
computations are hoisted, and it enables other passes like loop invariant
motion to remove more redundancies. The cost model to drive the code motion is
based on live range analysis on \SSA{} representation such that the (virtual)
register pressure does not increase.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
of 2.5\% savings in code size on \SPEC{}, although the code size also
increases in many cases because it enables more inlining. We have also seen
execution time improvements of approximately 5\% in a couple of \SPEC{} benchmarks viz. \GCC{}
and mcf, moreover, register spills reduced for almost all the \SPEC{}
bencmarks when compiled for X86\_64. This is an optimistic algorithm in the
sense that it considers all identical computations in a function as potential
candidates. We make an extra effort to hoist candidates by partitioning the
potential candidates in a way to enable partial hoisting in case common
hoisting points for all the candidates cannot be found. We also formalize how
register pressure will reduce as a result of code-motion and why sorting the
list of potential candidates w.r.t. their depth first numbers helps hoist more
candidates with less compile time overhead.
\end{abstract}

@@

Moreover, \LLVM{} has a `getelementptr' instruction which computes the address
where a load or a store would happen. It is a scalar computation and gets
hoisted frequently even if the corresponding loads/stores would not get hoisted.
In order to reduce register pressure while hoisting loads, we have restricted
hoisting of address computations away from their corresponding loads and stores
when the loads and stores cannot be moved.  This restriction is only to mitigate
the limitations of \LLVM{'}s register allocator and may be lifted in the future,
when the register allocation rematerialization pass has been improved to catch
these regressions.

@@

% TODO: Rewrite the following.
In our current implementation we keep as many candidates in one set as possible
(greedy approach). We split the list at a point where the legality checks fail
to hoist subset of candidates which are legal to hoist and then start finding
new hoisting point for the remaining ones.

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic block for all the loads. Even if the register pressure
would increase, we prefer to hoist loads in this case. The reason being, that
would make the loaded value available by the time it is used. Also, because
stores and calls are hoisted the least \ref{table:experimental-results}, the
performance does not change much whether they are hoisted or not.

@@

After one iteration of algorithm runs through the entire function, it creates
more oppportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the \GVN{} analysis to correctly populate congruence classes.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
We ran \SPEC{} with the patch. All
the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
optimization level.  The results for code-hoisting are listed in
Table~\ref{tab:code-motion-metric}. The table lists the number of scalars, loads,
stores and calls hoisted as well as removed. For each category, the number of
instructions removed is greater or equal to the number of instructions hoisted
because each hoisting is performed only when at least one identical computation
is found.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
Code size         &\% age	increase in codesize \\\hline
400.perlbench	  & 1.00 \\\hline
401.bzip2	  & 1.00 \\\hline
403.gcc	          & 1.01 \\\hline
429.mcf	          & 1.00 \\\hline
433.milc	  & 1.00 \\\hline
444.namd	  & 1.01 \\\hline
445.gobmk	  & 1.01 \\\hline
447.dealII	  & 1.01 \\\hline
450.soplex	  & 0.99 \\\hline
453.povray	  & 1.00 \\\hline
456.hmmer	  & 1.00 \\\hline
458.sjeng	  & 1.00 \\\hline
462.libquantum	  & 1.00 \\\hline
464.h264ref	  & 1.00 \\\hline
470.lbm	          & 1.00 \\\hline
471.omnetpp	  & 0.99 \\\hline
473.astar	  & 1.02 \\\hline
482.sphinx3	  & 1.00 \\\hline
483.xalancbmk	  & 0.99 \\\hline
Grand total       & 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in code size with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:code-size}
\end{table}

Code size remained almost the same as a result of code-motion as shown in Table~\cite{tab:code-size}.
The code size listed here is the text size (from linux size command) of the final executables for each benchmark.
Some bencmarks like 473.astar increased in code size by almost 2\%. This is because once the code-size
of a function decreases it becomes cheaper to inline. As we can see in Table-\cite{tab:stats}, four
more functions got inlined in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
all other benchmarks got more functions inlined.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
  Benchmark       &   Baseline	        &  GCM & \%Incrase     \\\hline
  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in compilation time with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:compile-time}
\end{table}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a global code motion compiler optimization which schedules congruent
instructions across the program.  Not only this code motion saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average of
2\% reduction in the number of spills, and saving of 1\% on overall compile time
on \SPEC{}. This being a code-size optimization, enables more inlining in
majority of the benchmarks. We have also seen execution
time improvements of approximately 5\% in a couple of \SPEC{} benchmarks
viz. \GCC{} and mcf, moreover, register spills reduced for almost all the
\SPEC{} bencmarks when compiled for X86\_64. This is an optimistic algorithm in
the sense that it considers all identical computations in a function as
potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of code-motion and why
sorting the list of potential candidates w.r.t. their depth first numbers helps
hoist more candidates with less compile time overhead.
\end{abstract}

@@

Moreover, \LLVM{} has a `getelementptr' instruction which computes the address
where a load or a store would happen. It is a scalar computation and gets
hoisted frequently even if the corresponding loads/stores would not be legal to
hoist.  In order to reduce the register pressure while hoisting loads, we have
restricted hoisting of address computations away from their corresponding loads
and stores when the loads and stores cannot be moved. This restriction is only
to mitigate the limitations of \LLVM{'}s register allocator and may be lifted in
the future, when the register allocation rematerialization pass has been
improved to catch these regressions.

@@

% TODO: Rewrite the following.
In our current implementation we keep as many candidates in one set as possible
(greedy approach.) We split the list at a point where the legality checks fail
to hoist subset of candidates which are legal to hoist and then start finding
new hoisting point for the remaining ones.

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic block for all the loads. Even if the register pressure
would increase, we prefer to hoist loads in this case. The reason being, that
would make the loaded value available by the time it is used. Also, because
stores and calls are hoisted the least, see Table~\ref{tab:code-motion-metric},
the performance does not change much whether they are hoisted or not.

@@

After one iteration of algorithm runs through the entire function, it creates
more oppportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the \GVN{} analysis to correctly populate congruence classes as
the program is modified by the code generation.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
We ran \SPEC{} with the patch. All
the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
optimization level.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
Benchmark       &\% change in performance \% change in codesize & \\\hline
400.perlbench	& 1.00  & 1.00 \\\hline
401.bzip2	& 1.00  & 1.00 \\\hline
403.gcc	        & 1.00  & 1.01 \\\hline
429.mcf	        & 1.03  & 1.00 \\\hline
433.milc	& 1.00  & 1.00 \\\hline
444.namd	& 1.00  & 1.01 \\\hline
445.gobmk	& 0.99  & 1.01 \\\hline
447.dealII	& 0.95  & 1.01 \\\hline
450.soplex	& 0.99  & 0.99 \\\hline
453.povray	& 0.97  & 1.00 \\\hline
456.hmmer	& 0.99  & 1.00 \\\hline
458.sjeng	& 1.02  & 1.00 \\\hline
462.libquantum	& 1.00  & 1.00 \\\hline
464.h264ref	& 1.00  & 1.00 \\\hline
470.lbm	        & 1.00  & 1.00 \\\hline
471.omnetpp	& 1.00  & 0.99 \\\hline
473.astar	& 1.00  & 1.02 \\\hline
482.sphinx3	& 0.99  & 1.00 \\\hline
483.xalancbmk	& 1.01  & 0.99 \\\hline
Geomean         & 0.997 & 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:code-size}
\end{table}

Overall there was only a minor change in performance regressed with a few ups
(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
etc.). The code-motion pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass and then after the inliner. Because EarlyCSE removes
local redundancies, GCM would have to analyze less redundant instructions, also
as inliner creates more redundancies, GCM would have more opportunities to move
congruent instructions out of branches.

Code size also remained almost the same as a result of code-motion as shown in
Table~\cite{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executables for each benchmark.  Some bencmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases it becomes cheaper to inline. As we can see in
Table-\cite{tab:stats}, four more functions got inlined in 473.astar. Except for
447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all other benchmarks got
more functions inlined.

  The results for code-hoisting are listed in
Table~\ref{tab:code-motion-metric}. The table lists the number of scalars, loads,
stores and calls hoisted as well as removed. For each category, the number of
instructions removed is greater or equal to the number of instructions hoisted
because each hoisting is performed only when at least one identical computation
is found.

@@

\begin{table}[h]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
  Benchmark       &   Baseline	        &  GCM & \%Increase     \\\hline
  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in compilation time with and without code-motion on SPEC2006 at -Ofast}
  \label{tab:compile-time}
\end{table}
